Colorimetric Detection of pH Strips

### 2017-12-10 CS189/paper/paper.tex
add and link references

\begin{abstract}
In this paper, we apply several machine learning techniques and models for the purpose of classifying and estimating the pH of solutions given raw image data of pH test strips. We replicate and critique much of the work performed by Mutlu et. al [1], who used LS-SVM and claimed to achieve 100\% classification. We believe that this high accuracy was the result of duplicating physical pH strips samples between training and validation datasets; this is problematic because the noise between different strips of the same class was not accounted for, and samples were essentially duplicated as pre-processing mitigated orientation variability efforts. We further find that regression is a more suitable approach for this domain, as pH values are on a continuous, logarithmic scale, and decimal differences in value can have significant biological consequences. In this spirit, we find that mean squared regression errors as low as $\sim$0.033 are achievable.
\end{abstract}

@@

\section{Introduction}
\IEEEPARstart{G}{rowing} cell cultures for industrial-scale chemical output is a very difficult task. Large bioreactors cost tens of thousands of dollars and require complex, specialized machinery to operate. An alternative idea is to make use of many 5-gallon buckets, with a film on top that allows air to pass through the system, but blocks liquids. With a pump inside the bucket to mix nutrients with the cells, we essentially have a low-cost bioreactor. We need to test the pH of each bucket periodically to determine when to feed the cells. This can be done in a cost-effective manner by using paper-based strips that are briefly submerged into the solution. This method of bioreactors can only be profitable if there are many buckets - on the order of hundreds or thousands. Therefore, we need an efficient, automated method of determining pH from the test strips. The literature on this suggest that in practice, smartphones are used as the source of these pictures.

Mutlu, et al., 2017, uses a Least Squares Support Vector Machine and a Support Vector Machine in order to classify images of pH strips by their pH values, capturing the images using a Smartphone camera and with various lighting and positional arrangement conditions [1].

@@

We collected our own data for this study. Mutlu was able to capture the range of pH values comprehensively through the use of adjustable chemical buffers. This allowed them to train on chemical solutions with integer pH values. They also were able to determine ground-truth labels for their data points by using a digitized pH meter. We were more limited in our resources; we had a fixed set of non-adjustable buffers and we took the label on the bottle as the ground-truth label for regression. In total we hand-held a smartphone to capture 10 samples under standard household lighting of pH strips submersed in the following 9 buffers: 2.22, 4.40, 4.47, 5.31, 6.28, 7.00, 7.70, 8.70, 10.0 (Fig. \ref{raw_photo}). These fixed non-integer pH values may be a source of irreducible error for us when we rounded these values to the nearest integer when performing classification. However, we wanted our model to be more robust to noise and to be applicable in a more reasonable environment, where a researcher would not have to use a custom apparatus remove any external sources of light [1, 2]. Mutlu waited for the strips to dry before taking pictures, but we decided to capture the images while they were still wet since we imagine that pH strips are read while wet in practice.

@@

From there, we extract the 12 features by taking the average value RBG values from each of the four color boxes (Fig. \ref{four-colors}a). We then generalize Mutlu’s experiments by artificially applying randomly generated lighting conditions to our data. This way, we do not correct for lighting before training a model like Mutlu does but rather, introduce lighting conditions to our homogeneous dataset so that our model can learn to ignore lighting conditions on its own, rather than having to rely on manual and lighting condition-specific pre-processing methods. Specifically, in our second experiment, we gamma correct each sample with a uniformly chosen coefficient between 0.5 and 1.5 (Fig. \ref{four-colors}b) because this mimics the nonlinear luminescence introduced in captured images [3]. In our third experiment, we perform hue rotation with a uniformly chosen degree between -90 and 90 degrees (Fig. \ref{four-colors}c). In our fourth experiment, we apply the random lighting conditions from both experiments 2 and 3 simultaneously (Fig. \ref{four-colors}d).

@@

The SVM classification model also performed quite well, and outperformed the SVM model implemented by Mutlu with respect to classification accuracy. In contrast with LDA, SVM attempts to find decision boundaries between the different classes of data. The color rotation in Experiment 3 added significant noise, so SVM performed better on these data with lower C, or stronger regularization, to combat this noise. On the other hand, the first two experiments involving the raw color data and the data with gamma modification achieved higher accuracies as C increased, or regularization became weaker. This makes sense because the colors in this data were closer to the ground truth color values for each class. In Experiment 4, SVM was unable to achieve an accuracy greater than $80\%$. Similarly, Mutlu experienced much lower classification accuracies around $60-80\%$ when applying SVMs to data with mixed light sources.

@@

The pH estimates from the best regression model tested had a very low mean squared error ($0.033$), making further research in the area of regression models for estimating pH very promising. Realistically, researchers working with pH test strips would likely prefer regression estimates for pH rather than rounded integer classification, as minute differences in pH can drastically alter the behavior of a solution. For example, in C2C12 cells, a commonly-used immortalized mouse myoblast cell line for mammalian muscular cell culture and research, the optimal growth pH is 7.0 to 7.6 [4]. If we do classification, a pH of 6.6, far too low for cell survival, will be classified to 7, while a pH of 7.6, which is in the optimal growth range, will be classified to 8. While integer binning was used to achieve $100\%$ classification accuracy in the Mutlu paper, these types of models may not be as useful as regression models in practice.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{1}

\bibitem{} A. Mutlu, V. Kılıç, G. Özdemir, A. Bayram, N. Horzum and M. Solmaz, \emph{Smartphone-based colorimetric detection via machine learning}, The Analyst, vol. 142, no. 13, pp. 2434-2441, 2017.
\bibitem{} S. Kim, Y. Koo and Y. Yun, \emph{A Smartphone-Based Automatic Measurement Method for Colorimetric pH Detection Using a Color Adaptation Algorithm}, Sensors, vol. 17, no. 7, p. 1604, 2017.
\bibitem{} H. Farid, \emph{Blind inverse gamma correction}, IEEE Transactions on Image Processing, vol. 10, no. 10, pp. 1428-1433, 2001.
\bibitem{} American Type Culture Collection, \emph{C2C12 (ATCC\textregistered CRL-177$^\texttt{TM}$)}, Product Sheet, Manassas, VA.

>>>>

\begin{abstract}
In this paper, we apply several machine learning techniques and models for the purpose of classifying and estimating the pH of solutions given raw image data of pH test strips. We replicate and critique much of the work performed by Mutlu et. al \cite{mutlu}, who used LS-SVM and claimed to achieve 100\% classification. We believe that this high accuracy was the result of duplicating physical pH strips samples between training and validation datasets; this is problematic because the noise between different strips of the same class was not accounted for, and samples were essentially duplicated as pre-processing mitigated orientation variability efforts. We further find that regression is a more suitable approach for this domain, as pH values are on a continuous, logarithmic scale, and decimal differences in value can have significant biological consequences. In this spirit, we find that mean squared regression errors as low as $\sim$0.033 are achievable.
\end{abstract}

@@

\section{Introduction}
\IEEEPARstart{G}{rowing} cell cultures for industrial-scale chemical output is a very difficult task. Large bioreactors cost tens of thousands of dollars and require complex, specialized machinery to operate. An alternative idea is to make use of many 5-gallon buckets, with a Tyvek membrane on top that allows air to pass through the system, but blocks liquids \cite{anderson}. With a pump inside the bucket to mix nutrients with the cells, we essentially have a low-cost bioreactor. We need to test the pH of each bucket periodically to determine when to feed the cells. This can be done in a cost-effective manner by using paper-based strips that are briefly submerged into the solution. This method of bioreactors can only be profitable if there are many buckets - on the order of hundreds or thousands. Therefore, we need an efficient, automated method of determining pH from the test strips. The literature on this suggest that in practice, smartphones are used as the source of these pictures \cite{mutlu}, \cite{kim}.

Mutlu, et al., 2017, uses a Least Squares Support Vector Machine and a Support Vector Machine in order to classify images of pH strips by their pH values, capturing the images using a Smartphone camera and with various lighting and positional arrangement conditions \cite{mutlu}.

@@

We collected our own data for this study. Mutlu was able to capture the range of pH values comprehensively through the use of adjustable chemical buffers. This allowed them to train on chemical solutions with integer pH values. They also were able to determine ground-truth labels for their data points by using a digitized pH meter. We were more limited in our resources; we had a fixed set of non-adjustable buffers and we took the label on the bottle as the ground-truth label for regression. In total we hand-held a smartphone to capture 10 samples under standard household lighting of pH strips submersed in the following 9 buffers: 2.22, 4.40, 4.47, 5.31, 6.28, 7.00, 7.70, 8.70, 10.0 (Fig. \ref{raw_photo}). These fixed non-integer pH values may be a source of irreducible error for us when we rounded these values to the nearest integer when performing classification. However, we wanted our model to be more robust to noise and to be applicable in a more reasonable environment, where a researcher would not have to use a custom apparatus remove any external sources of light \cite{mutlu}, \cite{kim}. Mutlu waited for the strips to dry before taking pictures, but we decided to capture the images while they were still wet since we imagine that pH strips are read while wet in practice.

@@

From there, we extract the 12 features by taking the average value RBG values from each of the four color boxes (Fig. \ref{four-colors}a). We then generalize Mutlu’s experiments by artificially applying randomly generated lighting conditions to our data. This way, we do not correct for lighting before training a model like Mutlu does but rather, introduce lighting conditions to our homogeneous dataset so that our model can learn to ignore lighting conditions on its own, rather than having to rely on manual and lighting condition-specific pre-processing methods. Specifically, in our second experiment, we gamma correct each sample with a uniformly chosen coefficient between 0.5 and 1.5 (Fig. \ref{four-colors}b) because this mimics the nonlinear luminescence introduced in captured images \cite{farid}. In our third experiment, we perform hue rotation with a uniformly chosen degree between -90 and 90 degrees (Fig. \ref{four-colors}c). In our fourth experiment, we apply the random lighting conditions from both experiments 2 and 3 simultaneously (Fig. \ref{four-colors}d).

@@

The SVM classification model also performed quite well, and outperformed the SVM model implemented by Mutlu with respect to classification accuracy. In contrast with LDA, SVM attempts to find decision boundaries by maximizing the margin between support vectors. The color rotation in Experiment 3 added significant noise, so SVM performed better on these data with lower C, or stronger regularization, to combat this noise. On the other hand, the first two experiments involving the raw color data and the data with gamma modification achieved higher accuracies as C increased, or regularization became weaker. This makes sense because the colors in this data were closer to the ground truth color values for each class. In Experiment 4, SVM was unable to achieve an accuracy greater than $80\%$. Similarly, Mutlu experienced much lower classification accuracies around $60-80\%$ when applying SVMs to data with mixed light sources.

@@

The pH estimates from the best regression model tested had a very low mean squared error ($0.033$), making further research in the area of regression models for estimating pH very promising. Realistically, researchers working with pH test strips would likely prefer regression estimates for pH rather than rounded integer classification, as minute differences in pH can drastically alter the behavior of a solution. For example, in C2C12 cells, a commonly-used immortalized mouse myoblast cell line for mammalian muscular cell culture and research, the optimal growth pH is 7.0 to 7.6 \cite{product}. If we do classification, a pH of 6.6, far too low for cell survival, will be classified to 7, while a pH of 7.6, which is in the optimal growth range, will be classified to 8. While integer binning was used to achieve $100\%$ classification accuracy in the Mutlu paper, these types of models may not be as useful as regression models in practice.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{1}
\bibitem{anderson} J. Anderson, \emph{2017\_10\_09-Final Project}, UC Berkeley, 2017.
\bibitem{mutlu} A. Mutlu, V. Kılıç, G. Özdemir, A. Bayram, N. Horzum and M. Solmaz, \emph{Smartphone-based colorimetric detection via machine learning}, The Analyst, vol. 142, no. 13, pp. 2434-2441, 2017.
\bibitem{kim} S. Kim, Y. Koo and Y. Yun, \emph{A Smartphone-Based Automatic Measurement Method for Colorimetric pH Detection Using a Color Adaptation Algorithm}, Sensors, vol. 17, no. 7, p. 1604, 2017.
\bibitem{farid} H. Farid, \emph{Blind inverse gamma correction}, IEEE Transactions on Image Processing, vol. 10, no. 10, pp. 1428-1433, 2001.
\bibitem{product} American Type Culture Collection, \emph{C2C12 (ATCC\textregistered CRL-177$^\texttt{TM}$)}, Product Sheet, Manassas, VA.
